Improvement of Thermostated Molecular Dynamics via Backward Error Analysis

### 2018-06-08 rigid_bodies.tex

\begin{tocentry}
	Graphical Abstract
\end{tocentry}

\begin{abstract}
	Abstract.
\end{abstract}

@@

Our new method is particularly appealing for rigid-body MD, primarily because its numerical stability extends up to time steps large enough to yield pronounced discretization errors. Rigid-body MD is relevant not only because some molecules are usually modeled as rigid, such as in the ubiquitous case of water \cite{Jorgensen_1983}, but also because it can provide a coarse-graining strategy for molecules that can be designed as collections of interconnected bodies, such as proteins and nanoparticles \cite{Miller_2002, Knorowski_2012, Patra_2013}.
In fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.

The case of rigid water exemplifies this behavior and serves as an excellent testbed for our proposal.
In fact, we are able to eliminate the discretization errors in numerically computed averages for time-step sizes up to 6 fs, outperforming both the double-chain method by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} and a reimplementation of the single-chain method by \citeauthor{Martyna_1996} \cite{Martyna_1996} for rigid bodies.

This paper proceeds as follows.
In Secs.~\ref{sec:nve}-\ref{sec:nvt} we review the rigid-body dynamics in the NVE and NVT ensembles.
This is followed by our results on backward error analysis, which we employ in Sec.~\ref{sec:refined_method} to derive our new NVT scheme, whose performance we assess in Sec.~\ref{sec:numerical_results}.
Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}. Appendix \ref{sec:rigid body shadow hamiltonian} contains a detailed derivation of the approximation to the shadow Hamiltonian.

@@

\subsection{NVE Dynamics: Notation and Symplectic Integration}
\label{sec:nve}
In a previous paper \cite{Silveira_2017}, we took advantage of a particular factorization of the orientation matrix expressed in terms of quaternion components to
1) derive a new exact solution for torque-free rotations and use it as part of a symplectic numerical integrator for the motion of rigid bodies and
2) reinterpret the approximate NO-SQUISH method \cite{Miller_2002}, thus unveiling its equivalence to a method previously developed by Dullweber and co-workers \cite{Dullweber_1997}.

The Hamiltonian system of ordinary differential equations (ODE) that describes the rigid-body motion is given by \cite{Silveira_2017}
\begin{subequations}
	\label{eq:ODE system for NVE}
	\begin{align}
%
	&\dot{\vt r} =
	%+\diff{\Ham{}}{\vt p} =
	{\mt M}^{-1} {\vt p}, \\
%
	&\dot{\vt p} =
	%-\diff{\Ham{}}{\vt r} = 
	{\vt F}, \\
%
	&\dot{\vt q} =
	%+\diff{\Ham{}}{\vt \pi} =
	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \label{eq:EDO_q} \\
%
	&\dot{\vt \pi} =
	%-\diff{\Ham{}}{\vt q} =
	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau, \label{eq:EDO_pi}
	\end{align}
\end{subequations}
In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the conjugate momenta of the quaternion.
Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference; and $\vt \omega$ $=$ $\frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$ is the three-dimensional angular velocity in the body-fixed frame.
Matrices $\mt M$ and $\mt I$ are diagonal.
The former contains the body mass in all three diagonal entries while the latter contains the three principal moments of inertia.
Finally, the matrix-valued functions $\mt B$ and $\vt C$ are defined as
\begin{equation*}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
%
\; \text{and} \;
%
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation*}

The factorization mentioned above, which played a central role in Ref.~\citenum{Silveira_2017}, is as simple as ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$, with ${\mt A}(\vt q)$ being the orientation matrix of the body (that is, the rotation matrix connecting the body-fixed and space-fixed frames).

@@

Of course, Eqs.~\eqref{eq:ODE system for NVE} to \eqref{eq:kinetic energy} can also represent a system of $N$ interacting rigid bodies if we consider that vectors $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$ contain the corresponding properties of all bodies piled together.
The same occurs for diagonal matrices $\mt M$ and $\mt I$, whose sizes increase to $3N \times 3N$ in this case.
In addition, $\mt B(\vt q)$ and $\mt C(\vt q)$ become block-diagonal matrices with $4N$ rows and $3N$ columns.
Yet, the notation can be made even more general if $\vt r$, $\vt p$, and $\mt M$ are enlarged further to accommodate a number of freely moving atoms.

@@

A symplectic, time-reversible integrator can be devised by splitting the exponential operator according to the Trotter-Suzuki formula \cite{Trotter_1959, Suzuki_1976}
\begin{equation}
\label{eq:splitting NVE}
e^{\timestep \Liu{\tiny NVE}} = e^{\frac{\timestep}{2} \Liu{B}} e^{\timestep \Liu{A}} e^{\frac{\timestep}{2} \Liu{B}} + \mathcal{O}(\timestep^3),
\end{equation}
where $\Liu{A} + \Liu{B} = \Liu{\tiny NVE}$.
In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick in the linear and quaternion momenta given by ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, respectively, where the asterisk denotes the state immediately before the propagation occurs.
The action of $e^{\timestep \Liu{A}}$ is the simultaneous uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) and torque-free rotation over a full time step. In Ref.~\citenum{Silveira_2017} we derived an exact solution of torque-free rotation in a form that facilitates the computer implementation allowing, differently to existing schemes \cite{Kosenko_1998,vanZon2007,Celledoni_2008}, a unified treatment for both asymmetric ($I_1$ $>$ $I_2$ $>$ $I_3$) and symmetric tops (either $I_1$ = $I_2$ or $I_2$ = $I_3$).
For mathematical details, the reader is referred to our previos work \cite{Silveira_2017}.
The exact solution comprise ``less familiar'' mathematical functions, such as the Jacobi elliptic functions and Carlson integrals.
On the other hand, the approximate solutions, such as the NO-SQUISH method  \cite{Miller_2002}, have proven to be very accurate in liquid-phase simulations \cite{Silveira_2017} and its formulation only involves sine and cosine terms.
This may explain why that the approximate solution is more commonly used \cite{vanZon2007}.
This pictures completely changes in the context of backward error analysis: in the approximate methods, an overall rotation is given by a sequence of uniform rotations around the principal axes of a body, which results in a factorization of the rotational operator given by $ e^{\frac{h}{2} i\!L_3} e^{\frac{h}{2} i\!L_2} e^{h i\!L_1} e^{\frac{h}{2} i\!L_2} e^{\frac{h}{2} i\!L_3}$, with each factor representing a rotation around a given axes.
The analytical solution, on the other hand, entails a single operator, thus making the application of backward error analysis considerably simpler.

>>>>

%\begin{tocentry}
%\end{tocentry}

%\begin{abstract}
%\end{abstract}

@@

Our new method is particularly appealing for rigid-body MD, primarily because its numerical stability extends up to time steps large enough to yield pronounced discretization errors. Rigid-body MD is relevant not only because some molecules are usually modeled as rigid, such as in the ubiquitous case of water \cite{Jorgensen_1983}, but also because it can provide a coarse-graining strategy for molecules that can be designed as collections of interconnected bodies, such as proteins and nanoparticles \cite{Miller_2002, Knorowski_2012, Patra_2013}.
In fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.
The case of rigid water exemplifies this behavior and serves as an excellent testbed for our proposal.
In fact, we were able to correct the discretization errors in numerically computed averages for time-step sizes up to 6 fs, outperforming both the double-chain method by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} and a reimplementation of the single-chain method by \citeauthor{Martyna_1996} \cite{Martyna_1996} for rigid bodies.

This paper proceeds as follows.
In Secs.~\ref{sec:nve}-\ref{sec:nvt} we review the rigid-body dynamics in the NVE and NVT ensembles.
This is followed by our developments on backward error analysis, which we employ in Sec.~\ref{sec:refined_method} to derive our new NVT scheme, whose performance we assess in Sec.~\ref{sec:numerical_results}.
Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}. Appendix \ref{sec:rigid body shadow hamiltonian} contains a detailed derivation of the shadow Hamiltonian approximation.

@@

\subsection{Hamiltonian Dynamics with Rigid Bodies: Notation and Symplectic Integration}
\label{sec:nve}

In a previous paper \cite{Silveira_2017}, we took advantage of a particular factorization of the orientation matrix expressed in terms of quaternion components to
1) derive a new exact solution for torque-free rotations and use it as part of a symplectic integrator for rigid bodies and
2) reinterpret the approximate NO-SQUISH method \cite{Miller_2002} and unveil its equivalence to the earlier method by Dullweber and co-workers \cite{Dullweber_1997}.

The Hamiltonian system of ordinary differential equations (ODE) that describes the rigid-body motion is given by \cite{Silveira_2017}
\begin{subequations}
	\label{eq:ODE system for NVE}
	\begin{align}
%
	&\dot{\vt r} =
	%+\diff{\Ham{}}{\vt p} =
	{\mt M}^{-1} {\vt p}, \\
%
	&\dot{\vt p} =
	%-\diff{\Ham{}}{\vt r} = 
	{\vt F}, \\
%
	&\dot{\vt q} =
	%+\diff{\Ham{}}{\vt \pi} =
	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \label{eq:EDO_q} \\
%
	&\dot{\vt \pi} =
	%-\diff{\Ham{}}{\vt q} =
	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau. \label{eq:EDO_pi}
	\end{align}
\end{subequations}

In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the quaternion-conjugate momentum.
Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference, and $\vt \omega$ $=$ $\frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$ is the three-dimensional angular velocity in the body-fixed frame.
Matrices $\mt M$ and $\mt I$ are diagonal ones.
The former contains the body mass in all three diagonal entries while the latter contains the three principal moments of inertia.
Finally, the matrix-valued functions $\mt B$ and $\vt C$ are defined as
\begin{equation*}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
%
\; \text{and} \;
%
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation*}

The factorization which played a central role in Ref.~\citenum{Silveira_2017} is ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$, where ${\mt A}(\vt q)$ is the rotation matrix that connects the body-fixed and space-fixed frames.

@@

Eqs.~\eqref{eq:ODE system for NVE} to \eqref{eq:kinetic energy} can also represent a system of $N$ interacting rigid bodies if we consider that vectors $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$ contain the corresponding properties of all bodies piled together.
The size of diagonal matrices $\mt M$ and $\mt I$ also increases to $3N \times 3N$ in this case.
In addition, $\mt B(\vt q)$ and $\mt C(\vt q)$ become block-diagonal matrices with $4N$ rows and $3N$ columns.
Yet, the notation can be made even more general if $\vt r$, $\vt p$, and $\mt M$ are enlarged further so as to accommodate a number of freely moving atoms.

@@

A symplectic, time-reversible integrator can be devised by splitting the exponential operator according to the Trotter-Suzuki formula \cite{Trotter_1959, Suzuki_1976}
\begin{equation}
\label{eq:splitting NVE}
e^{\timestep \Liu{\tiny NVE}} = e^{\frac{\timestep}{2} \Liu{B}} e^{\timestep \Liu{A}} e^{\frac{\timestep}{2} \Liu{B}} + \mathcal{O}(\timestep^3),
\end{equation}
where $\Liu{A} + \Liu{B} = \Liu{\tiny NVE}$.
In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick in the linear and quaternion momenta given by ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, respectively, where the asterisk denotes the state of the system immediately before the propagation occurs.
The action of $e^{\timestep \Liu{A}}$ is a simultaneous uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) and torque-free rotation along a full time step.
We have derived \cite{Silveira_2017} an exact solution for torque-free rotations in a form that facilitates computer implementation and allows, differently from other existing solutions \cite{Kosenko_1998, vanZon2007, Celledoni_2008}, a unified treatment of asymmetric, symmetric, and spherical tops.
This solution was used to evaluate the simpler NO-SQUISH method \cite{Miller_2002}.
Despite its approximate treatment of rotations, which are split into several revolutions around the principal axes \cite{Dullweber_1997}, that method proved to be very accurate in liquid-phase simulations \cite{Silveira_2017}.
This justifies its more widespread use.
However, having an unplit solution for rotations at hand is crucial for the feasibility of backward error analysis, as it might become clear in Sec.~\ref{sec:modified_h} and Appendix \ref{sec:rigid body shadow hamiltonian}.
